Starcode: sequence clustering based on all-pairs search

### 2014-11-06 doc/document.tex
Manuscript: corrected some typos.

\subsection{Benchmark}
\label{section:benchmark}
We benchmarked starcode against the sequence clustering algorithms
slidesort \citep{pmid21148542}, seed \citep{pmid21810899},
rainbow \citep{pmid22942077} and cd-hit \citep{pmid23060610}. Even though
slidesort is an all-pairs search algorithm, it was included in the
benchmark because sequence comparison is the most computationally
intensive step of the sequence clustering problem. Rainbow runs
exclusively on paired-end reads, whereas the other tools run on single
reads, for this reason all the tools could not be run no the same
dataset.

%% Synthetic data %%
The performance of sequence clustering algorithms can be sensitive
to the size of the clusters in the dataset, which in many applications
is not known \textit{a priori}. We therefore set up a benchmark on
artificial datasets to test the accuracy and the scaling of the
tools on a known cluster structure. We generated 4 datasets of 1 million
50-mers arranged in 1 to 1,000 clusters. Each cluster consisted
of 100 repeats of the same centroid sequence, plus satellites
derived from the centroid by incorporating 3 errors including at
most 1 indel. The number of satellites per cluster ranged from
999900 to 900. Rainbow was not tested on this benchmark because
the generated data is single reads. In addition, evaluating the
exactness of slidesort 
was problematic because the number of 3-matches in each dataset is
not known (pairs of statellites in the same cluster may be 3-matches
or not). For this reason we only compared the number of pairs found
by starcode with the number of pairs found by slidesort. The 
outcome of the test is summarized in
Figure \ref{fig:benchmark}.

@@

The running time of the different tools as a function of the size
of the clusters is shown on Figure \ref{fig:benchmark}c. The running
time of slidesort and starcode show a linear and sub-linear trends,
respectively. Seed and cd-hit run approximately in constant time
regardless of the cluster size. In spite of this result, the
performance of starcode remains competitive, even for clusters of 1
million sequences. The memory usage is shown in Figure
\ref{fig:benchmark}c. The smallest memory footprint is achieved by
slidesort and cd-hit, with a maximum difference of an order of
magnitude with respect to the other tools. Note that the comparison with
slidesort is not completely fair since it does not hold in memory the
full graph necessary for clustering (which is by far the most memory
demanding structure). The memory usage of starcode is the highest for
clusters of size 1000, but it decreases and becomes lower than the
memory usage of seed as the size of the cluster increases.

@@

The basic properties of the datasets used for benchmarking are
summarized in Table~\ref{tab:biobench_properties}. Dataset 1
(SRR950457) has been pre-processed to extract the barcode and remove
the constant part of the reads. Only barcodes between 15 and 17
nucleotides were included in the file. Dataset 2 (SRRHENCHI) consists
of raw Illumina single reads. These datasets differ by the read size,
the total read count and by the empirical cluster sizes. According to
the output of starcode, the largest clusters of dataset 1 contain
approximately 70,000 sequences, whereas dataset 2 contains 4 clusters
with more than 1 million sequences. Dataset 3 (SRR950477) has been
included to benchmark starcode against rainbow in paired-end
clustering mode.
\begin{table}[h]
\centering
\caption{Summary of the biological datasets used for benchmarking.
All the datasets are Illumina reads.}
\begin{tabular}{l|l c r}
    \textit{dataset}   & \textit{read count} &
    \textit{read length} & \textit{type} \\ \hline
    SRR950457 & 6,542,309    & 16     & single     \\
    SRRxxxxxx & 127,675,537  & 50     & single     \\
    SRR950477 & 2,460,226    & 50+50  & paired-end \\
\end{tabular}
\label{tab:biobench_properties}
\end{table}

The running times of starcode, seed, slidesort, rainbow and cd-hit
are summarized in Table~\ref{tab:biobench_runtime}. We accommodated the
distance threshold for the first dataset to compensate for the reduced
sequence length. Both starcode and slidesort were executed with
the option '-d 2' and the identity for cd-hit was set to '-c 0.85'. We
were not able to run seed on dataset 1 due to limitations on the
minimum input sequence length. Starcode was faster than the other
tools on all the datasets. CD-HIT and seed came in second position
with a running time approximately 35 and 20 times greater on datasets
1 and 2, respectively. Rainbow was nearly an order of magnitude slower
in the job of clustering paired-end reads. We did not record the exact
running times past 10 days since this is several orders of magnitude
higher than the running time of starcode.

@@

\begin{table}[h]
\centering
\caption{Memory usage (in GB).}
\begin{tabular}{l|c c c}
    \textit{software}   & \textit{dataset 1} &
      \textit{dataset 2} & \textit{dataset 3} \\ \hline
    starcode   & 0.65 & 30.9 & 5.2 \\
    seed       & -    & 53.9 & -   \\
    sidesort   & 1.30 & 13.9 & -   \\
    rainbow    & -    & -    & 0.5 \\
    cd-hit-est & 0.80 & 28.5 & -   \\
\end{tabular}
\label{tab:biobench_memory}
\end{table}

>>>>

\subsection{Benchmark}
\label{section:benchmark}
We benchmarked starcode against the sequence clustering algorithms
slidesort \citep{pmid21148542}, seed \citep{pmid21810899},
rainbow \citep{pmid22942077} and cd-hit \citep{pmid23060610}. Even though
slidesort is an all-pairs search algorithm, it was included in the
benchmark because sequence comparison is the most computationally
intensive step of the sequence clustering problem. Rainbow runs
exclusively on paired-end reads, whereas the other tools run on single
reads, for this reason all the tools could not be run on the same
dataset.

%% Synthetic data %%
The performance of sequence clustering algorithms can be sensitive
to the size of the clusters in the dataset, which in many applications
is not known \textit{a priori}. We therefore set up a benchmark on
artificial datasets to test the accuracy and the scaling of the
tools on a known cluster structure. We generated 4 datasets of 1 million
50-mers arranged in 1 to 1,000 clusters. Each cluster consisted
of 100 repeats of the same centroid sequence, plus satellites
derived from the centroid by incorporating 3 errors including at
most 1 indel. The number of satellites per cluster ranged from
999900 to 900. Rainbow was not tested on this benchmark because
the generated data are single reads. In addition, evaluating the
exactness of slidesort 
was problematic because the number of 3-matches in each dataset is
not known (pairs of statellites in the same cluster may be 3-matches
or not). For this reason we only compared the number of pairs found
by starcode with the number of pairs found by slidesort. The 
outcome of the test is summarized in
Figure \ref{fig:benchmark}.

@@

The running time of the different tools as a function of the size
of the clusters is shown on Figure \ref{fig:benchmark}c. The running
time of slidesort and starcode show linear and sub-linear trends,
respectively. Seed and cd-hit run approximately in constant time
regardless of the cluster size. In spite of this result, the
performance of starcode remains competitive, even for clusters of 1
million sequences. The memory usage is shown in Figure
\ref{fig:benchmark}c. The smallest memory footprint is achieved by
slidesort and cd-hit, with a maximum difference of an order of
magnitude with respect to the other tools. Note that the comparison with
slidesort is not completely fair since it does not hold in memory the
full graph necessary for clustering (which is by far the most memory
demanding structure). The memory usage of starcode is the highest for
clusters of size 1000, but it decreases and becomes lower than the
memory usage of seed as the size of the cluster increases.

@@

The basic properties of the datasets used for benchmarking are
summarized in Table~\ref{tab:biobench_properties}. Dataset 1
(SRR950457) has been pre-processed to extract the barcode and remove
the constant part of the reads. Only barcodes between 15 and 17
nucleotides were included in the file. Dataset 2 (SRRxxxxxx) consists
of raw Illumina single reads. These datasets differ by the read size,
the total read count and by the empirical cluster sizes. According to
the output of starcode, the largest clusters of dataset 1 contain
approximately 70,000 sequences, whereas dataset 2 contains 4 clusters
with more than 1 million sequences. Dataset 3 (SRR950477) has been
included to benchmark starcode against rainbow in paired-end
clustering mode.
\begin{table}[h]
\centering
\caption{Summary of the biological datasets used for benchmarking.
All the datasets are Illumina reads.}
\begin{tabular}{l|l c r}
    \textit{dataset}   & \textit{read count} &
    \textit{read length} & \textit{type} \\ \hline
    SRR950457 & 6,542,309    & 16     & single     \\
    SRRxxxxxx & 127,675,537  & 50     & single     \\
    SRR950477 & 2,460,226    & 50+50  & paired-end \\
\end{tabular}
\label{tab:biobench_properties}
\end{table}

The running times of starcode, seed, slidesort, rainbow and cd-hit
are summarized in Table~\ref{tab:biobench_runtime}. We accommodated the
distance threshold for the first dataset to compensate for the reduced
sequence length. Both starcode and slidesort were executed with
the option '-d 2' and the identity for cd-hit was set to '-c 0.85'. We
were not able to run seed on dataset 1 due to limitations on the
minimum sequence length. Starcode was significantly faster than the other
tools on all the datasets. CD-HIT and seed came in second position
with a running time approximately 35 and 20 times greater on datasets
1 and 2, respectively. Rainbow was nearly an order of magnitude slower
in the job of clustering paired-end reads. We did not record the exact
running times past 10 days since this is several orders of magnitude
higher than the running time of starcode.

@@

\begin{table}[h]
\centering
\caption{Memory usage (in GB).}
\begin{tabular}{l|c c c}
    \textit{software}   & \textit{SRR950457} &
      \textit{SRRxxxxxx} & \textit{SRR950477} \\ \hline
    starcode   & 0.65 & 30.9 & 5.2 \\
    seed       & -    & 53.9 & -   \\
    sidesort   & 1.30 & 13.9 & -   \\
    rainbow    & -    & -    & 0.5 \\
    cd-hit-est & 0.80 & 28.5 & -   \\
\end{tabular}
\label{tab:biobench_memory}
\end{table}
